---
title: "Cyclistic Bike-Share Analysis Case Study"
author: "Sweety Gain"
date: "2025-09-12"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Introduction

This analysis focuses on understanding how annual members and casual riders use Cyclistic bikes differently.
The goal is to provide data-driven insights and actionable recommendations to convert casual riders into annual members.


## Data Import and Cleaning 


### Load datasets
```{r}
tripdata_2019 <- read_csv("C:/Users/sweety Gain/OneDrive/Documents/Cyclistic_Case_Study/Divvy_Trips_2019_Q1.csv")
tripdata_2020 <- read_csv("C:/Users/sweety Gain/OneDrive/Documents/Cyclistic_Case_Study/Divvy_Trips_2020_Q1.csv")
```
### Combine datasets
```{r}
all_trips <- bind_rows(tripdata_2019, tripdata_2020)
```

### Data Cleaning
```{r}
all_trips <- all_trips %>%
  select(trip_id, started_at, ended_at, member_casual, rideable_type, start_station_name, end_station_name) %>%
  mutate(
    started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S"),
    ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S"),
    ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")),
    day_of_week = weekdays(started_at)
  ) %>%
  filter(
    !is.na(member_casual),
    ride_length > 0,
    ride_length <= 1440,
    ended_at <= as.POSIXct("2020-03-31 23:59:59")
  ) %>%
  select(-trip_id)
glimpse(all_trips)
```
### Avg length by user type
```{r}
avg_ride_length <- all_trips %>%
  group_by(member_casual) %>%
  summarize(avg_ride_length = mean(ride_length, na.rm = TRUE))

avg_ride_length
```
### Number of rides by day of the week
```{r}
rides_by_day <- all_trips %>%
  group_by(member_casual, day_of_week) %>%
  summarize(num_rides = n())
rides_by_day
```

### Visualization
# Average Ride Length by User Type
```{r}
ggplot(avg_ride_length, aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Ride Length by User Type",
       x = "User Type",
       y = "Average Ride Length (minutes)") +
  theme_minimal()
```

### Number of Rides by Day of the Week
```{r}
ggplot(rides_by_day, aes(x = day_of_week, y = num_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of Rides by Day of the Week",
       x = "Day of Week",
       y = "Number of Rides") +
  theme_minimal()
```

## Key Findings

- Annual members generally take shorter rides (average around 12 minutes) compared to casual riders (41 minutes).

- Casual riders predominantly use bikes on weekends, while members are active on weekdays, particularly during commuting hours (7–9 AM and 4–6 PM).

- Data shows a clear pattern that casual riders’ usage is leisure-focused, while members ride more for transportation needs.

## Top Three Recommendations

- Weekend Membership Promotions:
Design marketing campaigns specifically targeting casual riders who use bikes during weekends to encourage annual memberships.

- Digital Media Targeting:
Use digital ads highlighting cost benefits and convenience of annual membership, especially during peak leisure periods.

- Emphasize Commuter Benefits:
Develop campaigns that focus on the ease and savings for daily commuting when using an annual membership.